\section{Critical Reflection}
% ─────────────────────────────────────────────────────────────

This project has provided a challenging and illuminating end-to-end experience with applied machine learning—from raw multi-gigabyte network captures to production-oriented model comparison. The following reflection distils the most significant technical and epistemic lessons acquired.

\textbf{The Gap Between Theory and Practice.}
The formal guarantees of classical anomaly detectors—LOF's density consistency, OC-SVM's maximum-margin optimality—are formulated in idealised settings that frequently do not hold in practice. Our results starkly illustrate this: in our initial pipeline, LOF achieved a near-random AUC of 0.44. However, we discovered this was not merely an inherent limitation of high-dimensional spaces, but a consequence of a **methodological bug**: training a novelty-detection model (\texttt{novelty=True}) on a dataset contaminated with anomalies. Once we programmatically sanitised the pipeline to train LOF exclusively on clean normal traffic (preserving the algorithm's assumptions), its performance rose dramatically to an AUC of 0.76 (F1 = 0.53). This highlighted a key engineering lesson: selecting a model must be driven by both the geometry of the data and rigorous compliance with the algorithm's mathematical operational requirements.

\textbf{The Importance of Hyperparameter Sensitivity Analysis.}
The hyperparameter tuning exercise revealed that the contamination parameter is far more impactful than structural hyperparameters (such as the number of trees in IF or the number of neighbours in LOF). Small misspecifications of contamination—say, setting it to 0.20 when the true validation rate warrants a different threshold—can cost several percentage points of F1 on highly imbalanced data. This sensitivity is not prominently communicated in the original algorithm papers, which typically evaluate on balanced benchmarks. Systematic grid search followed by independent threshold sweeps on validation splits is therefore an essential engineering step rather than an optional refinement.

\textbf{On Evaluation Design.}
Perhaps the most important methodological lesson concerns evaluation. Standard accuracy on an 80/20 imbalanced dataset is almost entirely uninformative. A dummy classifier predicting ``benign'' for every flow achieves 80.3\% accuracy while catching zero attacks. Our three-perspective framework (original distribution, balanced subset, per-class analysis) was indispensable for exposing real model behaviour. The ensemble autoencoder's 97.1\% recall at 57.2\% precision illustrates the canonical cybersecurity trade-off: while a 42.8\% false alarm rate incurs human triage costs, catching nearly all attacks is paramount since the cost of a missed breach is potentially catastrophic. Designing evaluation protocols that reflect deployment objectives is a core competency that this project helped build.

\textbf{Ethical Dimensions.}
Anomaly detection systems do not operate in a vacuum. False positives can flag legitimate users as threats, with potential professional and access consequences, while false negatives can enable critical system breaches. Deploying such systems in sensitive infrastructure—healthcare networks or national utilities—demands absolute transparency about error rates and careful threshold calibration. Furthermore, since our models were trained on a Canadian university network capture from 2017 (CIC-IDS2017), their generalisation to other network topologies, cultural contexts, and modern evolving zero-day attacks cannot be assumed. Responsible deployment demands continuous drift monitoring, automated retraining, and explicit human oversight of high-confidence predictions.

\textbf{Broader Takeaways.}
This project successfully bridged the gap between classroom mathematical abstractions and the engineering realities of machine learning systems: handling OOM errors through stratified sampling, preventing data leakage by fitting standardizers exclusively on training splits, and balancing the computational feasibility of $\mathcal{O}(n^2)$ algorithms against their theoretical appeal. These engineering decisions are often invisible in academic literature but dominate real-world project timelines. Learning to make them thoughtfully—and to document the rationale—is, ultimately, as valuable as learning the algorithms themselves.
